---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

RGL - 3D visualization device system for R using OpenGL
=======================================================

```{r include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README"
)
options(rgl.useNULL=TRUE)
suppressPackageStartupMessages(library(rgl))

if (!requireNamespace("rmarkdown") || !rmarkdown::pandoc_available("1.14")) {
  warning(call. = FALSE, "These vignettes assume rmarkdown and pandoc version 1.14.  These were not found. Older versions will not work.")
  knitr::knit_exit()
}
setupKnitr(autoprint = TRUE)
```

```{r polyhedra, fig.height=1.5, echo = FALSE, snapshot = TRUE}
hcl.colors <- rgl:::hcl.colors  # might be cm.colors on old R
col <- hcl.colors(5)
mfrow3d(1,5)
shade3d(icosahedron3d(), col = col[1])
next3d()
shade3d(dodecahedron3d(), col = col[2])
next3d()
shade3d(octahedron3d(), col = col[3])
next3d()
shade3d(cube3d(), col = col[4])
next3d()
shade3d(tetrahedron3d(), col = col[5])
```

INTRODUCTION
------------
The RGL package is a visualization device system for R, using 
OpenGL or WebGL as the rendering backend. An OpenGL rgl device 
at its core is a real-time 3D engine written in C++. It provides an 
interactive viewpoint navigation facility (mouse + wheel support)
and an R programming interface.  WebGL, on the other hand, is
rendered in a web browser; rgl produces the input file, and the 
browser shows the images.

WEBSITE
-------

An experimental website is here:

 https://dmurdoch.github.io/rgl/dev

The currently active development site is here:

 https://github.com/dmurdoch/rgl


INSTALLATION
------------

Most users will want to install the latest CRAN release.
For Windows, macOS and some Linux platforms, installation can 
be easy, as CRAN distributes binary versions:

```
# Install latest release from CRAN
install.packages("rgl")
```

To install the latest development version from Github, 
you'll need to do a source install.  Those aren't easy!
Try

```
# Install development version from Github
remotes::install_github("dmurdoch/rgl")
```

If that fails, read the instructions below.


LICENSE
-------
The software is released under the GNU Public License. 
See [COPYING](./COPYING) for details.


FEATURES
--------
- portable R package using OpenGL (if available) on macOS, Win32 and X11
- can produce 3D graphics in web pages using WebGL
- R programming interface 
- interactive viewpoint navigation
- automatic data focus
- geometry primitives: 
  points, lines, triangles, quads, texts, point sprites
- high-level geometry:
  surface, spheres
- up to 8 light sources
- alpha-blending (transparency)
- side-dependent fill-mode rendering (dots, wired and filled)
- texture-mapping with mipmapping and environment mapping support
- environmental effects:
  fogging, background sphere
- bounding box with axis ticks marks
- undo operation:
  shapes and light-sources are managed on type stacks,
  where the top-most objects can be popped, or any item specified by
  an identifier can be removed


PLATFORMS
---------
macOS
Windows 7/10
Unix-derivatives


BUILD TOOLS
-----------
R recommended tools (gcc toolchain)
On Windows, Rtools40 (or earlier versions for pre-R-4.0.0)

REQUIREMENTS
------------
**For OpenGL display:**

Windowing System (unix/x11 or Windows)  
OpenGL Library  
OpenGL Utility Library (GLU)  

**For WebGL display:**

A browser with WebGL enabled.  See https://get.webgl.org.

Installing OpenGL support
-----------------------

**Debian:**  
aptitude install libgl1-mesa-dev libglu1-mesa-dev

**Fedora:**  
yum install mesa-libGL-devel mesa-libGLU-devel libpng-devel

**macOS:**  
Install XQuartz.  
`rgl` should work with either XQuartz 2.7.11 or 2.8.0, but
it will probably need rebuilding if the XQuartz version
changes.  XQuartz normally needs re-installation whenever the
macOS version changes.

**Windows:**  
Windows normally includes OpenGL support, but to get the
appropriate include files etc., you will need 
the appropriate version of
[Rtools](https://cran.r-project.org/bin/windows/Rtools/)
matched to your R version.  

Options
-------

The **libpng** library version 1.2.9 or newer is needed for pixmap import/export support.

The **freetype** library is needed for resizable 
anti-aliased fonts.  On Windows, it will be downloaded 
from https://github.com/rwinlib during the install.


BUILDING/INSTALLING
-------------------

Binary builds of `rgl` are available for some platforms on CRAN.

For source builds, install the prerequisites as described
above, download
the tarball and at the command line run

    R CMD INSTALL rgl_`r packageVersion("rgl")`.tar.gz

(with the appropriate version of the tarball).  The build
uses an `autoconf` configure script; to see the options, 
expand the tarball and run `./configure --help`.

Alternatively,
in R run

    install.packages("rgl")

to install from CRAN, or

    remotes::install_github("dmurdoch/rgl")
    
to install the development version from Github.


BUILDING WITHOUT OPENGL
-----------------------
  
As of version 0.104.1, it is possible to build the package without
OpenGL support on Unix-alikes (including macOS) with the configure option
    --disable-opengl 
For example,
  
    R CMD INSTALL --configure-args="--disable-opengl" rgl_`r packageVersion("rgl")`.tar.gz 
  
  On Windows, OpenGL support cannot currently be disabled.
  
  
DOCUMENTATION and DEMOS: 
------------------------

```
library(rgl)
browseVignettes("rgl")
demo(rgl)
```

CREDITS
-------
Daniel Adler <dadler@uni-goettingen.de>  
Duncan Murdoch <murdoch@stats.uwo.ca>  
Oleg Nenadic <onenadi@uni-goettingen.de>  
Simon Urbanek <simon.urbanek@math.uni-augsburg.de>  
Ming Chen <mchen34@uwo.ca>  
Albrecht Gebhardt <albrecht.gebhardt@uni-klu.ac.at>  
Ben Bolker <bolker@zoo.ufl.edu>  
Gabor Csardi <csardi@rmki.kfki.hu>  
Adam Strzelecki <ono@java.pl>  
Alexander Senger <senger@physik.hu-berlin.de>  
The R Core Team for some code from R.  
Dirk Eddelbuettel <edd@debian.org>  
The authors of Shiny for their private RNG code.  
The authors of knitr for their graphics inclusion code.
Jeroen Ooms for Rtools40 and FreeType help.  
Yohann Demont for Shiny code, suggestions, and testing.  
Joshua Ulrich for a lot of help with the Github migration.  
Xavier Fernandez i Marin for help debugging the build.

