---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

RGL - 3D visualization device system for R using OpenGL
=======================================================

```{r include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README"
)
options(rgl.useNULL=TRUE)
suppressPackageStartupMessages(library(rgl))

if (!requireNamespace("rmarkdown") || !rmarkdown::pandoc_available("1.14")) {
  warning(call. = FALSE, "These vignettes assume rmarkdown and pandoc version 1.14.  These were not found. Older versions will not work.")
  knitr::knit_exit()
}
setupKnitr(autoprint = TRUE)
```

```{r polyhedra, fig.height=1.5, echo = FALSE, snapshot = TRUE}
hcl.colors <- rgl:::hcl.colors  # might be cm.colors on old R
col <- hcl.colors(5)
mfrow3d(1,5)
shade3d(icosahedron3d(), col = col[1])
next3d()
shade3d(dodecahedron3d(), col = col[2])
next3d()
shade3d(octahedron3d(), col = col[3])
next3d()
shade3d(cube3d(), col = col[4])
next3d()
shade3d(tetrahedron3d(), col = col[5])
```

INTRODUCTION
------------
The RGL package is a visualization device system for R, using 
OpenGL or WebGL as the rendering backend. An OpenGL rgl device 
at its core is a real-time 3D engine written in C++. It provides an 
interactive viewpoint navigation facility (mouse + wheel support)
and an R programming interface.  WebGL, on the other hand, is
rendered in a web browser; rgl produces the input file, and the 
browser shows the images.


LICENSE
-------
The software is released under the GNU Public License. 
See "COPYING" for details.


FEATURES
--------
- portable R package using OpenGL (if available) on MacOSX, Win32 and X11
- can produce 3D graphics in web pages using WebGL
- R programming interface 
- interactive viewpoint navigation
- automatic data focus
- geometry primitives: 
  points, lines, triangles, quads, texts, point sprites
- high-level geometry:
  surface, spheres
- up to 8 light sources
- alpha-blending (transparency)
- side-dependant fill-mode rendering (dots, wired and filled)
- texture-mapping with mipmapping and environment mapping support
- environmental effects:
  fogging, background sphere
- bounding box with axis ticks marks
- undo operation:
  shapes and light-sources are managed on type stacks,
  where the top-most objects can be pop'ed, or any item specified by
  an identifier can be removed


PLATFORMS
---------
Mac OS X
Windows 7/10
Unix-derivatives


BUILD TOOLS
-----------
R recommended tools (gcc toolchain)
On Windows, Rtools40 (or earlier versions for pre-R-4.0.0)

REQUIREMENTS
------------
For OpenGL display:

Windowing System (unix/x11 or win32)
OpenGL Library 
OpenGL Utility Library (GLU)

For WebGL display:

Browser with WebGL enabled.

Installation on Debian:
-----------------------

To install OpenGL support:

aptitude install libgl1-mesa-dev libglu1-mesa-dev


OPTIONS
-------
libpng library version 1.2.9 or newer (pixmap import/export support)
freetype library (optional on Unix for resizable 
anti-aliased fonts), required on Windows


BUILDING/INSTALLING FROM SOURCE PACKAGE
---------------------------------------
The R build tool is the primary tool to build the RGL package.


BUILDING ON MICROSOFT WINDOWS
-----------------------------

Install Rtools40 or newer.

An Internet connection will be needed, as Freetype is
automatically downloaded from https://github.com/rwinlib.



BUILDING ON UNIX-STYLE OS (MacOS X, Linux, FreeBSD, ... ) 
---------------------------------------------------------

The build is controlled by an autoconf configure script.
You provide the options through the R CMD build/INSTALL command

  e.g. 
    $ R CMD INSTALL --configure-args="<configure args...>" rgl


  COMMON UNIX-STYLE OS OPTIONS
  ----------------------------

  --with-gl-includes=<path>
    GL C header files include path

  --with-gl-libraries=<path>
    GL library linkage path

  --with-gl-prefix=<libprefix>
    GL library prefix (e.g. Mesa)
  
  --disable-libpng
    disable libpng support

  --with-libpng-prefix=<install location>
    force LibPNG library install prefix (e.g. /usr/local)

  --disable-libpng-config
    explicitly disable libpng-config
  
  --disable-libpng-dynamic
    use static libpng library
    
  --disable-opengl
    disable all OpenGL displays; WebGL is still available
    
You may find that your distro doesn't have all necessary development
libraries installed:  read the error messages for hints!  This line
has been reported to be sufficient on FC 5 or 6:

 yum install mesa-libGL-devel mesa-libGLU-devel libpng-devel


  X11 WINDOWING SYSTEM OPTIONS
  ----------------------------

  The X11 windowing system is needed for OpenGL display in MacOS 
  and Unix-alikes.
  
  --x-includes=<path>
    X11 C header files include path

  --x-libraries=<path>
    X11 library linkage path


  BUILDING WITHOUT OPENGL
  -----------------------
  
  As of version 0.104.1, it is possible to build the package without
  OpenGL support on Unix-alikes (including MacOS) with the configure option
  --disable-opengl 
  For example,
  
  R CMD INSTALL --configure-args="--disable-opengl" rgl_0.104.1.tar.gz 
  
  On Windows, OpenGL support cannot currently be disabled.
  
  
DEMOS: LOADING AND RUNNING DEMONSTRATIONS
-------------------------------------------------------

```
library(rgl)
demo(rgl)
```

CREDITS
-------
Daniel Adler <dadler@uni-goettingen.de>  
Duncan Murdoch <murdoch@stats.uwo.ca>  
Oleg Nenadic <onenadi@uni-goettingen.de>  
Simon Urbanek <simon.urbanek@math.uni-augsburg.de>  
Ming Chen <mchen34@uwo.ca>  
Albrecht Gebhardt <albrecht.gebhardt@uni-klu.ac.at>  
Ben Bolker <bolker@zoo.ufl.edu>  
Gabor Csardi <csardi@rmki.kfki.hu>  
Adam Strzelecki <ono@java.pl>  
Alexander Senger <senger@physik.hu-berlin.de>  
The R Core Team for some code from R.  
Dirk Eddelbuettel <edd@debian.org>  
The authors of Shiny for their private RNG code.  
Jeroen Ooms for Rtools40 and Freetype help.  
Yohann Demont for Shiny code, suggestions, and testing.  
Joshua Ulrich for a lot of help with the Github migration.  

WEBSITE
-------
https://github.com/dmurdoch/rgl

