---
title: "Using rgl in pkgdown web sites"
author: "Duncan Murdoch"
output:
  rmarkdown::html_vignette:
    fig_height: 5
    fig_width: 5
    toc: yes
  pdf_document:
    fig_height: 5
    fig_width: 5
    toc: yes
  html_document:
    default
vignette: >
  %\VignetteIndexEntry{Using rgl in pkgdown web sites}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
if (!requireNamespace("rmarkdown", quietly = TRUE) ||
    !rmarkdown::pandoc_available("1.14")) {
  warning(call. = FALSE, "These vignettes assume rmarkdown and pandoc version 1.14.  These were not found. Older versions will not work.")
  knitr::knit_exit()
}
knitr::opts_chunk$set(echo = TRUE)
library(rgl)
options(rgl.useNULL = TRUE)
setupKnitr(autoprint = TRUE)
```


## What is the problem?

[pkgdown](https://pkgdown.r-lib.org/) is an R package that
makes it easy to build a web site for your package.  However, the current version 1.6.1 on CRAN doesn't work so well for
packages whose examples use `rgl` or other packages using
[htmlwidgets](http://htmlwidgets.org/) like
[leaflet](http://rstudio.github.io/leaflet).  This document
describes current progress in supporting both of these.

The main problem for [pkgdown](https://pkgdown.r-lib.org/) support
is that `rgl` and other [htmlwidgets](http://htmlwidgets.org/)
users require multiple Javascript libraries to be linked to
each web page.  `pkgdown` 1.6.1 can't do this, but I have
submitted 
[a pull request](https://github.com/r-lib/pkgdown/pull/1548)
that adds support for additional dependent libraries.  This 
also requires [changes to downlit](https://github.com/r-lib/downlit/pull/78), and
for other widgets, [changes to htmlwidgets](https://github.com/ramnathv/htmlwidgets/pull/404).

## Installing the changes

Installing the changes is fairly easy.  Make sure you have the
[remotes](https://remotes.r-lib.org/) package installed, then 
run

```{r eval=FALSE}
remotes::install_github(c("dmurdoch/downlit@rglpatch",
                          "dmurdoch/pkgdown@figures",
                          "dmurdoch/rgl",
                          "dmurdoch/htmlwidgets@rglpatch"))
```

With these development version packages installed, you should 
see `rgl` or `leaflet` output appear automatically in examples.
The `rgl` output is set up to mimic what would happen in a
`knitr` document that uses 
```{r eval=FALSE}
setupKnitr(autoprint = TRUE)
```
i.e. the output `rgl` commands will automatically be 
included in the display.  Low-level changes will be collected
into a single display:
```{r}
# Show regression plane with z as dependent variable

open3d()
x <- rnorm(100)
y <- rnorm(100)
z <- 0.2*x - 0.3*y + rnorm(100, sd = 0.3)
fit <- lm(z ~ x + y)

plot3d(x, y, z, type = "s", col = "red", size = 1)

# No plot here, because of the planes3d() call below

coefs <- coef(fit)
a <- coefs["x"]
b <- coefs["y"]
c <- -1
d <- coefs["(Intercept)"]
planes3d(a, b, c, d, alpha = 0.5)
```

## Is this safe to use?

These are development versions of the packages, so they may
contain bugs, either in my changes or in other unreleased 
changes.  I don't recommend you use them in regular production
code.  To re-install released versions of the packages, run
```{r eval=FALSE}
install.packages(c("downlit", "pkgdown", "htmlwidgets", "rgl"))
```

If you do use them and notice any bugs, please [report them](https://github.com/dmurdoch/rgl/issues)!
