---
title: "Using RGL in pkgdown web sites"
author: "Duncan Murdoch"
output:
  rmarkdown::html_vignette:
    fig_height: 5
    fig_width: 5
    toc: true
  pdf_document:
    fig_height: 5
    fig_width: 5
    toc: true
  html_document:
    default
vignette: >
  %\VignetteIndexEntry{Using RGL in pkgdown web sites}
  %\VignetteEngine{knitr::knitr}
---

```{r setup, include=FALSE}
if (!requireNamespace("rmarkdown", quietly = TRUE) ||
    !rmarkdown::pandoc_available("1.14")) {
  warning(call. = FALSE, "These vignettes assume rmarkdown and Pandoc
          version 1.14.  These were not found. Older versions will not work.")
  knitr::knit_exit()
}
knitr::opts_chunk$set(echo = TRUE, snapshot = FALSE, screenshot.force = FALSE)
library(rgl)
options(rgl.useNULL = TRUE)
setupKnitr(autoprint = TRUE)
```


## What is the problem?

[pkgdown](https://pkgdown.r-lib.org/) is an R package that
makes it easy to build a web site for your package.  However, the previous version 1.6.1 on CRAN didn't work so well for
packages whose examples use RGL or other packages like [leaflet](http://rstudio.github.io/leaflet/) that use
[htmlwidgets](http://www.htmlwidgets.org).  This document
described changes that support both of these.  The changes
are now incorporated into `pkgdown` 2.0.0, so everything but
user instructions has been removed.

## Using `RGL in examples.

If you use RGL in example code, they should
appear automatically in examples.
The RGL output is set up to mimic what would happen in a
`knitr` document that uses 
```{r eval=FALSE}
setupKnitr(autoprint = TRUE)
```
i.e. the output RGL commands will automatically be 
included in the display.  Low-level changes will be collected
into a single display:
```{r}
# Show regression plane with z as dependent variable
library(rgl)
open3d()
x <- rnorm(100)
y <- rnorm(100)
z <- 0.2*x - 0.3*y + rnorm(100, sd = 0.3)
fit <- lm(z ~ x + y)

plot3d(x, y, z, type = "s", col = "red", size = 1)

# No plot here, because of the planes3d() call below

coefs <- coef(fit)
a <- coefs["x"]
b <- coefs["y"]
c <- -1
d <- coefs["(Intercept)"]
planes3d(a, b, c, d, alpha = 0.5)
```

## Specifying the size of figures

By default, pkgdown generates standard plots which are wider
than they are high (according to the golden ratio).  Often
RGL plots look better with equal width and height, since
the contents may be rotated by the user.

To specify the size of plots in pkgdown, you use the `figures`
entry in `_pkgdown.yml`.  The defaults are similar to
```
figures:
  dev: ragg::agg_png
  dpi: 96
  dev.args: []
  fig.ext: png
  fig.width: 7
  fig.height: ~
  fig.retina: 2
  fig.asp: 1.618
  bg: NA
  other.parameters: []
```

By default RGL uses these parameters as well, but allows you 
to override any of `fig.width`, `fig.height` and `fig.asp` by
specifying an `rgl` entry in `other.parameters`.  For example:
```
figures:
  fig.width: 5
  other.parameters:
    rgl:
      fig.asp: 1
```
This will make all plots have a width of 5 inches
and will make RGL plots square.
