---
title: "Rendering Text"
author: "Duncan Murdoch"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Rendering Text}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

These are notes about the internal details of rendering text
in `rgl`.

# General notes

Whether rendering single glyphs or strings, the following
concepts apply.

- the "baseline" is the reference line all letters sit on.
Letters with "descenders" go below the baseline.

- the "actual bounding box" covers the leftmost to rightmost
and topmost to bottommost pixels that will be drawn.

The actual bounding box is used when packing characters 
into a texture.  It is generally safest to expand the actual bounding box by at least one pixel on each side, so that 
anti-aliasing doesn't respond to other nearby pixels.


- the "reference point" is the point specified for the
start of text when drawing.  It's generally on the baseline.  
Sometimes it will be different (in either direction) from 
the leftmost side of the actual bounding box.

- the "advance" is the amount by which the imagined cursor
should advance to position another glyph or string after the
current one.

- the "font bounding box" runs horizontally from the reference point to that point plus the advance. The verticals run from
the baseline up to the ascent of the font (essentially the 
topmost location for any character in
the font, but fonts are allowed to write outside these bounds).

The font bounding box is used to set positioning of the text.
The convention in `rgl` is similar to base graphics:  vertical 
position 0 is the baseline, vertical position 1 is the top of
the font bounding box.  Horizontal position 0 is the left side
and position 1 is the right side of that box.

These plots show font bounding boxes for strings
rendered in the Times New Roman italic font.  In each case the
baseline is at $y=0$ and the reference point is $(0,0)$.

```{r echo = FALSE}
library(rgl)
plotBox <- function(text, family = "Times New Roman", 
                    font = 3, cex = 5, 
                    box = c("font", "actual"), 
                    xlab = "", ylab = "",
                    col = if (box == "font") "red" else "blue"
                    ) {
  box <- match.arg(box)
  atlas <- glyphAtlas(text, family=family, font=font, 
                      cex = cex)
  xlim <- c()
  ylim <- c()
  s <- atlas$strings
  if (box == "font") {
    bottom <- 0
    left <- 0
    right <- left + atlas$strings[1, "width"]
    top <- bottom + s[1, "height"]
    ylim <- range(c(bottom, top))
  }
  for (i in seq_len(nrow(atlas$fragments))) {
    fragment <- atlas$fragments[i,]
    glyphs <- cbind(fragment, atlas$glyphs[fragment$glyphnum,])
    xlim <- range(c(xlim, 
                    glyphs$x_offset + glyphs$x,
                    glyphs$x_offset + glyphs$x + glyphs$width))
    ylim <- range(c(ylim,
                   glyphs$y_offset - glyphs$y,
                   glyphs$y_offset - glyphs$y - glyphs$height))
  }
  if (box == "actual") {
    left <- xlim[1]
    right <- xlim[2]
    bottom <- ylim[1]
    top <- ylim[2]
  }
  ylim <- range(c(ylim, 0,  s[1, "height"]))
  renderFromAtlas(atlas, 1,
                  ylim = ylim, xlab = xlab, ylab = ylab)
  segments(c(left, right, right, left),
          c(bottom, bottom, top, top),
          c(right, right, left, left),
          c(bottom, top, top, bottom), col=col)
}
plotBox("Mf")
plotBox("Âg")
```

The actual bounding box is used internally to set the limits
of the textures holding the strings.
These plots show actual bounding boxes:

```{r echo=FALSE}
plotBox("Mf", box="actual")
plotBox("Âg", box="actual")
```

# Implementation in `rgl` master branch

The master branch of `rgl` currently uses the `FTGL` library
to render text.  It uses Freetype to render to a texture
and then displays the text using the texture.

See the FTGL source for implementation details.

# Implementation in `rgl` shaders branch

The shaders branch (to become `rgl` version 1.4.x) implements
text drawing internally in C++ code.

Currently it uses `Pango`, `Cairo` and `freetype` for
rendering and font handling, and
`harfbuzz` for shaping.  This may change before release.

An atlas of glyphs (characters) is maintained; see the
declarations in
`src/atlas.h` for the parts.  The R function `glyphAtlas`
provides an interface to this code, mainly used for debugging.
Its man page gives an overview of the layout of the atlas.

For rendering, each character is drawn separately, at positions
determined by `harfbuzz` and stored in the glyph atlas 
structure.

The function `pango_font_metrics_get_ascent()` is used to
determine the top of the font bounding box and it is stored
for string `i` as `Glyph_atlas.strings[i].height` (in R, `atlas$strings$height[i]`).

The function `pango_font_get_glyph_extents()` is used to 
determine the size and positioning of each glyph; these
are stored in `Glyph_atlas.glyphs` (in R, `atlas$glyphs`). 
The width and height are rounded
up to the next whole pixel, but fractional positioning is
kept.  Positioning in the texture is also stored here.

The function `hb_buffer_get_glyph_positions()` is used to
determine positioning of each glyph in the output; the results
are stored in `atlas$fragments`.

The atlas stores information about its current state in
`atlas$position`; this is updated as new glyphs are added.

This is the code used to set the texture coordinates and 
adjustment offsets:
```
      // Set texcoords
      texCoordArray[4*g].s = (glyph.x_atlas)/texture_width;
      texCoordArray[4*g + 1].s = texCoordArray[4*g].s + glyph.width/texture_width;
      texCoordArray[4*g + 2].s = texCoordArray[4*g + 1].s;
      texCoordArray[4*g + 3].s = texCoordArray[4*g].s;

      double texture_height = atlas.height;
      texCoordArray[4*g].t = (glyph.y_atlas + glyph.height)/texture_height;
      texCoordArray[4*g + 1].t = texCoordArray[4*g].t;
      texCoordArray[4*g + 2].t = texCoordArray[4*g].t - glyph.height/texture_height;
      texCoordArray[4*g + 3].t = texCoordArray[4*g + 2].t;
      
      // Set adjustments
      adjArray.setVertex(4*g,
        Vertex((-adj.x*s.width + s.x_offset[j] + glyph.x)*rescale,
                (-adj.y*s.height + s.y_offset[j] - glyph.y - glyph.height)*rescale,
                0) );      
      adjArray.setVertex(4*g + 1,
        Vertex(adjArray[4*g].x + glyph.width*rescale,
               adjArray[4*g].y,
               0));
      adjArray.setVertex(4*g + 2,
        Vertex(adjArray[4*g + 1].x,
               adjArray[4*g].y + glyph.height*rescale,
               0));
      adjArray.setVertex(4*g + 3,
        Vertex(adjArray[4*g].x,
               adjArray[4*g + 2].y,
               0));
```

# Implementation in WebGL

The WebGL implementation used by `rglwidget()` is written
in Javascript.  Currently it doesn't use the glyph atlas
described above, it stores whole strings in the texture.

This code uses the `CanvasRenderingContext2D: measureText()` 
function to measure text, but doesn't use it for height, where
it bases the height of the font bounding box on twenty times
the `cex` value.  The actual bounding box width is used for 
the font bounding box width.  (Prior to `rgl` 1.3.38 the font
bounding box width was used where the actual width should have
been used, which led to the clipping described in issue #503).

Eventually these calculations should be made
consistent with the ones used in the main display shaders branch.

This is the code used to set the texture coordinates and 
offsets in the shaders branch of the rglwidget Javascript code:
```
      rescale = fl.fixed_size ? 2 : 0.013;
      for (i=0; i < v.length; i++) {
        adj = this.getAdj(obj, i, offset, obj.texts[i]);
        vnew[i]  = v[i].concat([0,0]).concat(adj);
        fnew[4*i] = f[i];
        vnew[last] = vnew[i].slice();
        fnew[4*i+1] = last++;
        vnew[last] = vnew[i].slice();
        fnew[4*i+2] = last++;
        vnew[last] = vnew[i].slice();
        fnew[4*i+3] = last++;
        alias[i] = [last-3, last-2, last-1];
        v1 = vnew[fnew[4*i]];
        v1[oofs] = (-adj[0]*texinfo.fwidths[i] - texinfo.offsetX[i])*rescale;
        v1[oofs+1] = (-adj[1]*texinfo.fheights[i] 
                      - texinfo.heights[i] + texinfo.offsetY[i]
                      )*rescale;
        v1[oofs+2] = 0;
        v1[tofs] = texinfo.canvasX[i]/texinfo.canvasWidth;
        v1[tofs+1] = 1 - (texinfo.canvasY[i] + texinfo.heights[i])/texinfo.canvasHeight;
        vnew[fnew[4*i]] = v1; // Not necessary?
        
        v2 = v1.slice();
        v2[oofs] = v2[oofs] + texinfo.widths[i]*rescale;
        v2[tofs] = v1[tofs] + texinfo.widths[i]/texinfo.canvasWidth;
        vnew[fnew[4*i+1]] = v2;
        
        v3 = v2.slice();
        v3[oofs+1] = v3[oofs+1] + texinfo.heights[i]*rescale;
        v3[tofs+1] = v1[tofs+1] + texinfo.heights[i]/texinfo.canvasHeight;
        vnew[fnew[4*i+2]] = v3;
        
        v4 = v1.slice();
        v4[oofs+1] = v3[oofs+1];
        v4[tofs+1] = v3[tofs+1];
        vnew[fnew[4*i+3]] = v4;
      }
      v = vnew;
      obj.vertexCount = v.length;
      obj.f = [fnew, fnew];
```
The master branch WebGL code pays less attention to actual 
text measurements.