---
title: "WebGL"
author: "Duncan Murdoch"
date: "January 26, 2015"
output: html_document
---


```{r setup, echo=FALSE, results="asis"}
source("setup.R")
set.seed(123)
```

This document describes how to use embedded Javascript to 
modify a WebGL document.

For example, consider the simple plot of the iris data.  We
insert a code chunk with label `plot3d` (which will be used below).
```{r plot3d, rgl=TRUE}
with(iris, plot3d(Sepal.Length, Sepal.Width, Petal.Length, 
                  type="s", col=as.numeric(Species)))
currentSubscene3d()
```
We might like a button on the web page to cause a change to the
display, e.g. a rotation of the plot.  First we add buttons, with
the "onclick" event set to a function described below:

    <button type="button" onclick="rotate(10)">Forward</button>
    <button type="button" onclick="rotate(-10)">Backward</button>

which produces these buttons: 
<button type="button" onclick="rotate(10)">Forward</button>
<button type="button" onclick="rotate(-10)">Backward</button>

We found the subscene number that is currently active using 
`currentSubscene3d()` in the code chunk above.

The `rotate()` function makes use of the global `<prefix>rgl`
object.  The `knitr` WebGL support sets the prefix to the 
code chunk label, so the global is called `plot3drgl`:

    <script type="text/javascript">
    var rotate = function(angle) {
      plot3drgl.userMatrix[1].rotate(angle, 0,1,0);
      plot3drgl.drawScene();
    }
    </script>
    
<script type="text/javascript">
var rotate = function(angle) {
  plot3drgl.userMatrix[1].rotate(angle, 0,1,0);
  plot3drgl.drawScene();
}
</script>

We can also change the contents of the plot using a button.
For example, we can redo the previous plot, but with the
three species as separate "spheres" objects:
```{r toggle, rgl=TRUE}
with(subset(iris, Species == "setosa"), 
     spheres3d(Sepal.Length, Sepal.Width, Petal.Length, 
                  col=as.numeric(Species),
                  radius = 0.211))
with(subset(iris, Species == "versicolor"), 
     spheres3d(Sepal.Length, Sepal.Width, Petal.Length, 
               col=as.numeric(Species),
     	       radius = 0.211))
with(subset(iris, Species == "virginica"), 
     spheres3d(Sepal.Length, Sepal.Width, Petal.Length, 
               col=as.numeric(Species),
     	       radius = 0.211))
aspect3d(1,1,1)
decorate3d()
currentSubscene3d()
rgl.ids()
```
Now we can add buttons to turn the species on and off:

<button type="button" onclick="toggle(1)">Setosa</button>
<button type="button" onclick="toggle(2)">Versicolor</button>
<button type="button" onclick="toggle(3)">Virginica</button>
<script type="text/javascript">
var toggle = function(species) {
  id = species + 24;
  var objects = togglergl.opaque[19];
  var i = objects.indexOf(id);
  if (i == -1)
    objects.push(id);
  else
    objects.splice(i, 1);
  togglergl.opaque[19] = objects;
  togglergl.drawScene();
}
</script>

This used the following code:
```
<button type="button" onclick="toggle(1)">Setosa</button>
<button type="button" onclick="toggle(2)">Versicolor</button>
<button type="button" onclick="toggle(3)">Virginica</button>
<script type="text/javascript">
var toggle = function(species) {
  id = species + 24;
  var objects = togglergl.opaque[19];
  var i = objects.indexOf(id);
  if (i == -1)
    objects.push(id);
  else
    objects.splice(i, 1);
  togglergl.opaque[19] = objects;
  togglergl.drawScene();
}
</script>
```
We used the following here:

- Our code chunk had label `toggle`, so we use the `togglergl` global.
- The magic number 24 is chosen because the spheres for each species
are 25, 26, and 27, as `rgl.ids` showed above. 
- The magic number 19 is the subscene number.
- `togglergl.opaque[19]` gives an array that lists all the opaque
objects in subscene 19.  Also available are `togglergl.transparent[19]`
and `togglergl.subscenes[19]`.
