---
title: "WebGL"
author: "Duncan Murdoch"
date: "January 26, 2015"
output: html_document
---


```{r setup, echo=FALSE, results="asis"}
source("setup.R")
set.seed(123)
```

This document describes how to use embedded Javascript to 
modify a WebGL document.

For example, consider the simple plot of the iris data.  We
insert a code chunk with label `plot3d` (which will be used below).
```{r plot3d, rgl=TRUE}
with(iris, plot3d(Sepal.Length, Sepal.Width, Petal.Length, 
                  type="s", col=as.numeric(Species)))
currentSubscene3d()
```
We might like a button on the web page to cause a change to the
display, e.g. a rotation of the plot.  First we add buttons, with
the "onclick" event set to a function described below:

    <button type="button" onclick="rotate(10)">Forward</button>
    <button type="button" onclick="rotate(-10)">Backward</button>

which produces these buttons: 
<button type="button" onclick="rotate(10)">Forward</button>
<button type="button" onclick="rotate(-10)">Backward</button>

We found the subscene number that is currently active using 
`currentSubscene3d()` in the code chunk above.

The `rotate()` function makes use of the global `<prefix>rgl`
object.  The `knitr` WebGL support sets the prefix to the 
code chunk label, so the global is called `plot3drgl`:

    <script type="text/javascript">
    var rotate = function(angle) {
      plot3drgl.userMatrix[1].rotate(angle, 0,1,0);
      plot3drgl.drawScene();
    }
    </script>
    
<script type="text/javascript">
var rotate = function(angle) {
  plot3drgl.userMatrix[1].rotate(angle, 0,1,0);
  plot3drgl.drawScene();
}
</script>