---
title: "User Interaction in WebGL"
author: "Duncan Murdoch"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  rmarkdown::html_vignette:
    toc: yes
    fig_width: 5
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{User Interaction in WebGL} 
  %\VignetteEngine{knitr::rmarkdown}
---


```{r setup, echo=FALSE, results="asis"}
source("setup.R")
set.seed(123)
```

## Introduction

This document describes how to use embedded Javascript to 
control a WebGL display in an HTML document.  For more 
general information, see [rgl Overview](rgl.html).

We start with two simple examples.  The next section gives 
reference information.

Consider the simple plot of the iris data.  We
insert a code chunk with label `plot3d` (which will be used below).
```{r plot3d, rgl=TRUE}
with(iris, plot3d(Sepal.Length, Sepal.Width, Petal.Length, 
                  type="s", col=as.numeric(Species)))
subid <- currentSubscene3d()
```

We might like a button on the web page to cause a change to the
display, e.g. a rotation of the plot.  First we add buttons, with
the "onclick" event set to a function described below:

    <button type="button" onclick="rotate(10)">Forward</button>
    <button type="button" onclick="rotate(-10)">Backward</button>

which produces these buttons: 
<button type="button" onclick="rotate(10)">Forward</button>
<button type="button" onclick="rotate(-10)">Backward</button>

We stored the subscene number that is currently active in
`subid` in the code chunk above, and use it as `r rinline("subid")`
in the script below.

The `rotate()` function makes use of the global `<prefix>rgl`
object.  The `knitr` WebGL support sets the prefix to the 
code chunk label, so the global is called `plot3drgl`:

    <script type="text/javascript">
    var rotate = function(angle) {
      plot3drgl.userMatrix[`r rinline("subid",
                           script=TRUE)`].rotate(angle, 0,1,0);
      plot3drgl.drawScene();
    }
    </script>
    
<script type="text/javascript">
var rotate = function(angle) {
  plot3drgl.userMatrix[`r subid`].rotate(angle, 0,1,0);
  plot3drgl.drawScene();
}
</script>

## Autogenerated controls

We can also change the contents of the plot using `r indexfns("toggleButton")`.
For example, we can redo the previous plot, but with the
three species as separate "spheres" objects and buttons to
toggle them:
```{r toggle, rgl = TRUE, rgl.keepopen = TRUE}
sphereid <- with(subset(iris, Species == "setosa"), 
     spheres3d(Sepal.Length, Sepal.Width, Petal.Length, 
                  col=as.numeric(Species),
                  radius = 0.211))
with(subset(iris, Species == "versicolor"), 
     spheres3d(Sepal.Length, Sepal.Width, Petal.Length, 
               col=as.numeric(Species),
     	       radius = 0.211))
with(subset(iris, Species == "virginica"), 
     spheres3d(Sepal.Length, Sepal.Width, Petal.Length, 
               col=as.numeric(Species),
     	       radius = 0.211))
aspect3d(1,1,1)
decorate3d()
subid <- currentSubscene3d()
```

```{r results="asis"}
toggleButton(sphereid, label = "setosa", prefix = "toggle", subscene = subid)
toggleButton(sphereid+1, label = "versicolor", prefix = "toggle", subscene = subid)
toggleButton(sphereid+2, label = "virginica", prefix = "toggle", subscene = subid)
```

Note that we need to use `results="asis"` for the button code.  I didn't
use it above, but normally we would use `echo=FALSE`; then the buttons 
will end up side-by-side:

```{r results="asis", echo=FALSE}
toggleButton(sphereid, "setosa", "toggle", subid)
toggleButton(sphereid+1, "versicolor", "toggle", subid)
toggleButton(sphereid+2, "virginica", "toggle", subid)
```

An alternate control to achieve the same thing is `r indexfns("subsetSlider")`. 

```{r slider, rgl=TRUE}
```
```{r results="asis"}
subsetSlider(list(setosa = sphereid, 
		  versicolor = sphereid + 1, 
		  virginica = sphereid + 2, 
		  all = sphereid + 0:2),
	     prefix = "slider", subscene = subid)
```

Another function that auto-generates Javascript code is 
`r indexfns("clipplaneSlider")`.  This function allows the user to control
the location of a clipping plane by moving a slider.  

## User Defined Mouse Controls

`rgl` allows user defined mouse controls.  For these to work
within WebGL, you will need to write a Javascript version as
well as the R version.

For example, the help page for `r linkfn("rgl.setMouseCallbacks")` defines
a `pan3d` function to set a callback for panning a scene:
```{r echo=FALSE, results = "hide"}
 open3d(useNULL=TRUE)
```

```
 pan3d <- function(button) {
   start <- list()
   
   begin <- function(x, y) {
       start$userMatrix <<- par3d("userMatrix")
       start$viewport <<- par3d("viewport")
       start$scale <<- par3d("scale")
       start$projection <<- rgl.projection()
       start$pos <<- rgl.window2user( x/start$viewport[3], 1 - y/start$viewport[4], 0.5, 
                                      projection = start$projection)
   }
   
   update <- function(x, y) {
        xlat <- (rgl.window2user( x/start$viewport[3], 1 - y/start$viewport[4], 0.5,
                                 projection = start$projection) - start$pos)*start$scale
        mouseMatrix <- translationMatrix(xlat[1], xlat[2], xlat[3])
        par3d(userMatrix = start$userMatrix %*% t(mouseMatrix) )
   }
   rgl.setMouseCallbacks(button, begin, update)
   cat("Callbacks set on button", button, "of rgl device", rgl.cur(), "\n")
 }
 pan3d(3)
```
<script type="text/javascript">
var start = new Object();
var begin = function(x,y) {
  
}
</script>
TODO:  DESCRIBE HOW TO WRITE THIS IN JAVASCRIPT

## Reference for `rglClass`

In writing the `writeWebGL()` function, I haven't tried to prevent access to 
anything.  On the other hand, I haven't provided access to
everything.  The parts documented here should remain relatively stable
(unless indicated otherwise).  Users may also consult the source
to `writeWebGL`, but should be aware that anything that isn't documented
here is subject to change without notice.  

`r indexclass("rglClass")`

As documented in `r linkfn("writeWebGL")`, the call
```{r eval=FALSE}
writeWebGL(..., prefix = "<prefix>")
```
will create a global object on the output page with name 
`<prefix>rgl` and Javascript class `rglClass`. 
This class has a large number of properties and methods, some of which are designed
to be available for use by other code on the web page. 

Most of the properties are stored as Javascript `Array` objects, indexed
by the `rgl` id of the subscene to which they apply.  There 
are also Javascript methods attached to the `rglClass` class.

### Methods

`r indexmethods("drawScene")`

After any change that will affect the display, code should
call `<prefix>rgl.drawScene()` to redraw the scene.

`r indexmethods(c("inSubscene", "addToSubscene", "delFromSubscene"))`

These methods each take two arguments:  `id` and `subscene`,
which should be the `rgl` ids of an object and a subscene.
`inSubscene` tests whether `id` is already included in the 
subscene, and the others 
add it or delete it from the subscene.

`r indexmethods(c("getSubsceneEntries"))`

This function takes a subscene id as argument, and returns an `Array`
containing all of the ids displayed in that subscene.

`r indexmethods(c("setSubsceneEntries"))`

This takes an `Array` of ids and a subscene id as arguments, and sets
the contents of the subscene to the ids.

### Properties

`r indexproperties(c("FOV", "listeners", "userMatrix", "zoom"))`

These correspond to the 
`r linkfn("par3d")` properties with the same names.  

- `FOV` and `zoom` are arrays of numbers. 
- `userMatrix` is an array
of `CanvasMatrix4` objects (documented in the file 
`system.file("WebGL/CanvasMatrix.src.js")`.
- Each `listeners` item is itself an array of subscene ids that "listen"
to mouse actions, i.e. `listeners[19]` would contain all
subscene ids that respond to mouse actions in subscene 19.

`r indexproperties("viewport")`

This property also corresponds to the
`r linkfn("par3d")` property, but should be considered to be 
read-only.

`r indexproperties(c("drawFns", "clipFns"))`

These two arrays contain the code to display 
each object in the scene.  The functions in the
`drawFns` array are called for each object
each time it is displayed.  The `clipFns` functions
are called when objects being clipped are drawn.

## Index 

The following functions and `rglClass` properties and methods are described in this document:<br>

```{r echo=FALSE, results="asis"}
writeIndex(cols = 5)
```


