---
title: "WebGL"
author: "Duncan Murdoch"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  rmarkdown::html_vignette:
    toc: yes
    fig_width: 5
    fig_height: 5
vignette: >
  %\VignetteIndexEntry{WebGL} 
  %\VignetteEngine{knitr::rmarkdown}
---


```{r setup, echo=FALSE, results="asis"}
source("setup.R")
set.seed(123)
```

## Introduction

This document describes how to use embedded Javascript to 
modify a WebGL document.
We start with two simple examples.  The next section gives 
reference information.

Consider the simple plot of the iris data.  We
insert a code chunk with label `plot3d` (which will be used below).
```{r plot3d, rgl=TRUE}
with(iris, plot3d(Sepal.Length, Sepal.Width, Petal.Length, 
                  type="s", col=as.numeric(Species)))
currentSubscene3d()
```
We might like a button on the web page to cause a change to the
display, e.g. a rotation of the plot.  First we add buttons, with
the "onclick" event set to a function described below:

    <button type="button" onclick="rotate(10)">Forward</button>
    <button type="button" onclick="rotate(-10)">Backward</button>

which produces these buttons: 
<button type="button" onclick="rotate(10)">Forward</button>
<button type="button" onclick="rotate(-10)">Backward</button>

We found the subscene number that is currently active using 
`currentSubscene3d()` in the code chunk above.

The `rotate()` function makes use of the global `<prefix>rgl`
object.  The `knitr` WebGL support sets the prefix to the 
code chunk label, so the global is called `plot3drgl`:

    <script type="text/javascript">
    var rotate = function(angle) {
      plot3drgl.userMatrix[1].rotate(angle, 0,1,0);
      plot3drgl.drawScene();
    }
    </script>
    
<script type="text/javascript">
var rotate = function(angle) {
  plot3drgl.userMatrix[1].rotate(angle, 0,1,0);
  plot3drgl.drawScene();
}
</script>

We can also change the contents of the plot using a button.
For example, we can redo the previous plot, but with the
three species as separate "spheres" objects:
```{r toggle, rgl=TRUE}
with(subset(iris, Species == "setosa"), 
     spheres3d(Sepal.Length, Sepal.Width, Petal.Length, 
                  col=as.numeric(Species),
                  radius = 0.211))
with(subset(iris, Species == "versicolor"), 
     spheres3d(Sepal.Length, Sepal.Width, Petal.Length, 
               col=as.numeric(Species),
     	       radius = 0.211))
with(subset(iris, Species == "virginica"), 
     spheres3d(Sepal.Length, Sepal.Width, Petal.Length, 
               col=as.numeric(Species),
     	       radius = 0.211))
aspect3d(1,1,1)
decorate3d()
currentSubscene3d()
rgl.ids()
```
Now we can add buttons to turn the species on and off:

<button type="button" onclick="toggle(1)">Setosa</button>
<button type="button" onclick="toggle(2)">Versicolor</button>
<button type="button" onclick="toggle(3)">Virginica</button>
<script type="text/javascript">
var toggle = function(species) {
  id = species + 24;
  var objects = togglergl.opaque[19];
  var i = objects.indexOf(id);
  if (i == -1)
    objects.push(id);
  else
    objects.splice(i, 1);
  togglergl.opaque[19] = objects;
  togglergl.drawScene();
}
</script>

This used the following code:
```
<button type="button" onclick="toggle(1)">Setosa</button>
<button type="button" onclick="toggle(2)">Versicolor</button>
<button type="button" onclick="toggle(3)">Virginica</button>
<script type="text/javascript">
var toggle = function(species) {
  id = species + 24;
  var objects = togglergl.opaque[19];
  var i = objects.indexOf(id);
  if (i == -1)
    objects.push(id);
  else
    objects.splice(i, 1);
  togglergl.opaque[19] = objects;
  togglergl.drawScene();
}
</script>
```
We used the following here:

- Our code chunk had label `toggle`, so we use the `togglergl` global.
- The magic number 24 is chosen because the spheres for each species
are 25, 26, and 27, as `rgl.ids` showed above. 
- The magic number 19 is the subscene number.
- `togglergl.opaque[19]` gives an array that lists all the opaque
objects in subscene 19.  Also available are `togglergl.clipplanes[19]`,
`togglergl.transparent[19]` and `togglergl.subscenes[19]`.

## Reference Information

In writing the `writeWebGL()` function, I haven't tried to prevent access to 
anything.  On the other hand, I haven't provided access to
everything.  The parts documented here should remain relatively stable
(unless indicated otherwise).  Users may also consult the source
to `writeWebGL`, but should be aware that anything that isn't documented
here is subject to change without notice.  

As documented in `r linkfn("writeWebGL")`, the call
```{r eval=FALSE}
writeWebGL(..., prefix = "<prefix>")
```
will create a global object on the output page with name `<prefix>rgl`.  
This object has a large number of properties, some of which are designed
to be available for modification by other code on the web page. 

Most of these properties are stored as Javascript `Array` objects, indexed
by the `rgl` id of the subscene to which they apply.

`r indexprop(c("FOV", "listeners", "userMatrix", "zoom"))`

These correspond to the 
`r linkfn("par3d")` properties with the same names.  

- `zoom` and `FOV` are arrays of numbers. 
- `userMatrix` is an array
of `CanvasMatrix4` objects (documented in the file 
`system.file("WebGL/CanvasMatrix.src.js")`.
- Each `listeners` item is itself an array of subscene ids that "listen"
to mouse actions, i.e. `listeners[19]` would contain all
subscene ids that respond to mouse actions in subscene 19.

`r indexprop("viewport")`

This property also corresponds to the
`r linkfn("par3d")` property, but should be considered to be 
read-only.

`r indexprop(c("clipplanes", "opaque", "transparent", "subscenes"))`

Entries in these are arrays of object ids to 
display in the current subscene.  Objects need
to be displayed in the given order so should not
be moved between the arrays, but
as in the "toggle" example above, objects can be removed
from these lists to affect whether they are displayed or 
not.  

Objects should also not be moved between 
subscenes:  to simulate that, include the objects
in all subscenes in the R code, then remove 
extras in Javascript code run when the web page loads. 

`r indexprop(c("drawFns", "clipFns"))`

These two arrays contain the code to display 
each object in the scene.  The functions in the
`drawFns` array are called for each object
each time it is displayed.  The `clipFns` functions
are called when objects being clipped are drawn.

## Index of Properties

The following `rglClass` properties are described in this document:<br>

```{r echo=FALSE, results="asis"}
writeIndex(cols = 5)
```


